Experience Report: Haskell in Computational Biology

### 2012-06-27 mrfy_experience_report.tex
good libraries save "effort" not "time
edit section 2 and start section 3

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell: 
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
but the implementation required little effort,
and the connection between code and
algorithm is unusually clear (\secref{viterbi}).
Our new Haskell code performs competitively with~C++,
and it was easy to parallelize
(\secref{perf}). 
\remark{Can we summarize the performance results?  Speedup?}
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one way: it's hard to understand and
improve the
performance of a Haskell program (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although authors Daniels and Gallant are computational
biologists with little background in functional programming,
we~used Haskell successfully.
And Haskell has made it easy to explore new research ideas;
by~contrast,
our group's C++ code has not been a good vehicle for exploring new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some worked as advertised, saved us lots of time, and were a pleasure
to~use.
Others, not so much.
And~we could not tell ahead of time which experience we were likely to have
 (\secref{penumbra}).
\end{itemize}

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signaling, and so on.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
One of the best known methods of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given a {query sequence} of amino acids,
find known proteins of similar structure.
When similar proteins are formed from amino-acid sequences that
are not too different from the query sequence, the homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research contribution---a software tool called MRFy (pronounced
``Murphy'')---can detect homologies 
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%

@@

Homology-detection software is most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
Either way, 
the software is \emph{trained}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into a correspondence, also called an
\emph{alignment}. 
This alignment relates individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins (Figure~\ref{alignment}).

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
We~represent the state chosen for each node using an algebraic data type:
\remark{we promised a better name for HMMState. Thoughts?}
\begin{smallverbatim}
data HMMState = Mat | Ins | Del | Beg | End
\end{smallverbatim}

The particular hidden Markov model we use requires that every nonempty
sequence of insertions be followed by a match, and similarly for
deletions.
In~other words, insertions and deletions cannot be mixed without an
intervening match.
This restriction means that there are only seven possible transitions
into the states of any one column \cite{Eddy:1998ut}. 
\begin{itemize} 
\item
A~transition into a match state in
column~$j$ is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or $\txprobj D M$.
\item
A~transition into a deletion state in
 column~$j$ is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$ or $\txprobj D D$.
\item
A~transition into an insertion state in
 column~$j$ is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$ or $\txprobjj I I$.
\end{itemize}

@@

The log of a probability is just a number, which we call a \texttt{Score}.
A transition probability connects two \texttt{HMMState}s, and so we wrap it in
a value constructor:
\begin{smallverbatim}
newtype Score = Score Double

data TProb =
     TProb { logProbability :: Score
           , fromState :: HMMState
           , toState :: HMMState
           }

>>>>

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
When performance is paramount,
software is usually written in C~or~C++. 
When convenience, readability, and
productivity are more important,
software is usually written in a dynamically typed
or domain-specific language like
Perl, Python, Ruby, SPSS, or~R.
In~this paper, we report on experience using a third kind of language,
Haskell: 
\begin{itemize}
\item
We had to reimplement an
algorithm already implemented in~C++, 
but the implementation required little effort,
and the connection between code and
algorithm is unusually clear (\secref{viterbi}).
Our new Haskell code performs competitively with~C++,
and it was easy to parallelize
(\secref{perf}). 
\remark{Can we summarize the performance results?  Speedup?}
\item
Higher-order functions made it unusually easy to
create and experiment with new stochastic-search algorithms
(\secref{hofs}).
\item
Haskell slowed us down in only one way: it's hard to understand and
improve the
performance of a Haskell program (\secref{awkward-profiling}).
%%  \item
%%  The only significant impediment that Haskell presented to our
%%  research was the difficulty of understanding and improving the
%%  performance of Haskell programs (\secref{awkward-profiling}).
\item
Although authors Daniels and Gallant are computational
biologists with little background in functional programming,
we~used Haskell successfully.
And Haskell has made it easy to explore new research ideas;
by~contrast,
our group's C++ code has not been a good vehicle for exploring new ideas
(\secref{comparo}).
\item
The Haskell community offers libraries and tools that
promise powerful abstractions.
Some worked as advertised, saved us lots of effort, and were a pleasure
to~use.
Others, not so much.
And~we could not tell ahead of time which experience we were likely to have
 (\secref{penumbra}).
\end{itemize}

@@

Proteins %%% are cellular machinery. They
interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signaling, and so on.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence, biologists wish to know the cellular
function of the protein that gene codes for.
One of the best known methods of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are expected to be
descended from a common ancestor---in biological terms, \emph{homologous}---and
thus
the problem of identifying proteins similar to a \textit{query sequence} is called 
\textit{homology detection}.

Computational biologists detect homologies by building 
algorithms which, given a {query sequence}, %% of amino acids,
find known proteins of similar structure.
When similar proteins are formed from amino-acid sequences that
are not too different from the query sequence, the homology can be
detected by
a family of algorithms called 
\textit{hidden Markov models}~\cite{Eddy:1998ut}.
But in real biological systems,
proteins with similar structure and function may be formed from significantly 
different amino-acid sequences, which are not close in edit distance.
Our~research software, MRFy (pronounced
``Murphy''), can detect homologies 
in amino-acid sequences that are only distantly related.
%
%We will explain an 
%algorithm for detecting reasonably similar sequences, and then move on to 
%explain MRFy, our novel approach for detecting more distantly related sequences 
%for proteins that share similar structure and function.
%

@@

Homology-detection software is most often used in one of two ways:
to test a hypothesis about 
the function of a single, newly discovered protein, or 
to compare every protein in a genome against a library of known protein 
structures.
Either way, 
the software is \emph{trained}
on a group of proteins that share function and structure.
These proteins are identified by a biologist, who puts
their amino-acid sequences into an
\emph{alignment}. 
This alignment relates individual amino acids in a set of homologous proteins.
An~alignment may be represented as a matrix
in which each row corresponds to the amino-acid sequence of a protein,
and each column groups amino acids that play similar roles in
different proteins (Figure~\ref{alignment}).

@@

A~hidden Markov model carries probabilities on some states and on all
state transitions.
Both the probabilities and the states are determined by the alignment:
\remark{These bullets need proper parallel structure}
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.
In~our representation, each state contains a probability or a table of
probabilities, and it is also labelled with one of these labels:
\smallverbatiminput{statelabel}

We~use the ``Plan7'' hidden Markov model, which forbids direct
transitions between insertion 
states and deletion states \cite{Eddy:1998ut}. 
The model gets its name because there are exactly
seven possible transitions
into the states of any column~$j$.
Each transition has its own probability:
\begin{itemize} 
\item
A~transition into a match state 
is more likely when column~$j$ is a consensus column.
Depending on the predecessor state, 
the probability of such a transition is 
$\txprobj M M$, $\txprobj I M$, or~$\txprobj D M$.
\item
A~transition into a deletion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobj M D$~or~$\txprobj D D$.
\item
A~transition into an insertion state 
is more likely when column~$j$ is a non-consensus column.
The probability of such a transition is 
$\txprobjj M I$~or~$\txprobjj I I$.
\end{itemize}

@@

The log of a probability is just a number, which we call a \texttt{Score}.
A transition probability connects two \texttt{StateLabel}s, and so we wrap it in
a value constructor:
\begin{smallverbatim}
newtype Score = Score Double

data TProb =
     TProb { logProbability :: Score
           , fromState :: StateLabel
           , toState :: StateLabel
           }
